\documentclass[11pt]{extbook}

\usepackage{geometry}

\usepackage{fancyvrb}
\usepackage{color}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage{verbatim}
\usepackage{tikz}
\usepackage{listings}
\usepackage[yyyymmdd,hhmmss]{datetime}
\usepackage{rotating}
\usepackage{authblk}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{todonotes}
\usepackage{titlesec}
\usepackage[mathlines]{lineno}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{bm}
\usepackage{etoolbox}
\usepackage{pdflscape}
\usepackage{threeparttable}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{graphicx}

%TGM:  Added these packages to fix underscore rendering
\usepackage{lmodern} 
\usepackage[T1]{fontenc}

\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}

\titleformat{\paragraph}
{\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
\titlespacing*{\paragraph}
{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}

\newtoggle{assign}
\toggletrue{assign}

\newcommand{\qg}{\u{g}}
\newcommand{\qG}{\u{G}}
\newcommand{\qc}{\c{c} }
\newcommand{\qC}{\c{C}}
\newcommand{\qs}{\c{s}}
\newcommand{\qS}{\c{S}}
\newcommand{\qu}{\"{u}}
\newcommand{\qU}{\"{U}}
\newcommand{\qo}{\"{o}}
\newcommand{\qO}{\"{O}}
\newcommand{\qI}{\.{I}}
\newcommand{\wa}{\^{a}}
\newcommand{\wA}{\^{A}}

\begin{document}

\linenumbers

\newcommand{\kron}{\mathbin{\text{\footnotesize \textcircled{\raisebox{-0.3pt}{\footnotesize $\otimes$}}}}}
\newcommand{\grbarray}[1]{\bm{#1}}
\newcommand{\scalar}[1]{{#1}}
\renewcommand{\vector}[1]{{\bf #1}}
\renewcommand{\matrix}[1]{{\bf #1}}
\renewcommand{\arg}[1]{{\sf #1}}
\newcommand{\zip}{{\mbox{zip}}}
\newcommand{\zap}{{\mbox{zap}}}
\newcommand{\ewiseadd}{{\mbox{\bf ewiseadd}}}
\newcommand{\ewisemult}{{\mbox{\bf ewisemult}}}
\newcommand{\mxm}{{\mbox{\bf mxm}}}
\newcommand{\vxm}{{\mbox{\bf vxm}}}
\newcommand{\mxv}{{\mbox{\bf mxv}}}
\newcommand{\gpit}[1]{{\sf #1}}
\newcommand{\ie}{{i.e.}}
\newcommand{\eg}{{e.g.}}
\newcommand{\nan}{{\sf NaN}}
\newcommand{\nil}{{\bf nil}}
\newcommand{\ifif}{{\bf if}}
\newcommand{\ifthen}{{\bf then}}
\newcommand{\ifelse}{{\bf else}}
\newcommand{\ifendif}{{\bf endif}}
\newcommand{\zero}{{\bf 0}}
\newcommand{\one}{{\bf 1}}
\newcommand{\true}{{\sf true}}
\newcommand{\false}{{\sf false}}
\newcommand{\syntax}{{Math Notation}}

\newcommand{\Dinn}{\mbox{$D_{in}$}}
\newcommand{\Din}[1]{\mbox{$D_{in_{#1}}$}}
\newcommand{\Dout}{\mbox{$D_{out}$}}

\newcommand{\bDinn}{\mbox{$\mathbf{D}_{in}$}}
\newcommand{\bDin}[1]{\mbox{$\mathbf{D}_{in_{#1}}$}}
\newcommand{\bDout}{\mbox{$\mathbf{D}_{out}$}}

\newcommand{\scott}[1]{{{\color{violet}[Scott: #1]}}}
\newcommand{\tim}[1]{{{\color{teal}[Tim: #1]}}}
\newcommand{\ben}[1]{{{\color{blue}[Ben: #1]}}}
\newcommand{\will}[1]{{{\color{red}[Will: #1]}}}

%\newcommand{\scott}[1]{}
%\newcommand{\tim}[1]{}
%\newcommand{\ben}[1]{}
%\newcommand{\will}[1]{}

\renewcommand{\comment}[1]{{}}
\newcommand{\glossBegin}{\begin{itemize}}
\newcommand{\glossItem}[1]{\item \emph{#1}: }
\newcommand{\glossEnd}{\end{itemize}}

\setlength{\parskip}{0.5\baselineskip}
\setlength{\parindent}{0ex}

%\usepackage{draftwatermark}
%\SetWatermarkText{DRAFT}
%\SetWatermarkScale{2}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\setcounter{footnote}{1}

\input{graphblas_math_commands}

%-----------------------------------------------------------------------------

\title{
The GraphBLAS Math Specification
\footnote{Based on \emph{GraphBLAS Mathematics} by Jeremy Kepner}: \\ 
{\large Version 1.0.0 (or 2.0?} \\
{\normalsize \scott{THIS IS A DRAFT VERION. Update acks and remove DRAFT before release.}}
}

\author{Benjamin Brock, Ayd\i n Bulu\c{c}, Timothy Mattson, Scott McMillan, Jos\'e Moreira}

\date{Generated on \today\ at \currenttime\ EDT}

\maketitle


\renewcommand{\thefootnote}{\arabic{footnote}}
\setcounter{footnote}{0}

\vfill

Copyright \copyright\ 2022 Carnegie Mellon University, The Regents 
of the University of California, through Lawrence Berkeley National 
Laboratory (subject to receipt of any required approvals from the 
U.S. Dept. of Energy), the Regents of the University of California 
(U.C. Berkeley), Intel Corporation, International Business Machines 
Corporation, and Massachusetts Institute of Technology Lincoln
Laboratory. 

Any opinions, findings and conclusions or recommendations expressed in 
this material are those of the author(s) and do not necessarily reflect 
the views of the United States Department of Defense, the United States 
Department of Energy, Carnegie Mellon University, the Regents of the 
University of California, Intel Corporation, or the IBM Corporation.  

NO WARRANTY. THIS MATERIAL IS FURNISHED ON AN AS-IS BASIS. THE COPYRIGHT 
OWNERS AND/OR AUTHORS MAKE NO WARRANTIES OF ANY KIND, EITHER EXPRESSED 
OR IMPLIED, AS TO ANY MATTER INCLUDING, BUT NOT LIMITED TO, WARRANTY OF 
FITNESS FOR PURPOSE OR MERCHANTABILITY, EXCLUSIVITY, OR RESULTS OBTAINED 
FROM USE OF THE MATERIAL. THE COPYRIGHT OWNERS AND/OR AUTHORS DO NOT MAKE 
ANY WARRANTY OF ANY KIND WITH RESPECT TO FREEDOM FROM PATENT, TRADE MARK, 
OR COPYRIGHT INFRINGEMENT.

\vspace{1.5cm}

\vspace{2cm}
%{\Large This version is a definitive release of the GraphBLAS C API
%specification. As of the date of this document, at least two independent
%and functionally complete implementations are available.}

%{\Large This version is a provisional release of the GraphBLAS C API specification.
%Once two functionally complete reference implementations are available, we
%will remove the "provisional" clause.}


\vspace{1.5cm}


%[Distribution Statement A] This material has been approved for public release and unlimited distribution.  
%Please see Copyright notice for non-US Government use and distribution.

Except as otherwise noted, this material is licensed under a Creative Commons Attribution 4.0 license (\href{http://creativecommons.org/licenses/by/4.0/legalcode}{http://creativecommons.org/licenses/by/4.0/legalcode}), 
and examples are licensed under the BSD License (\href{https://opensource.org/licenses/BSD-3-Clause}{https://opensource.org/licenses/BSD-3-Clause}).

%\begin{abstract}
%\end{abstract}

\vfill

\pagebreak
\tableofcontents
\vfill
\pagebreak

%-----------------------------------------------------------------------------

\phantomsection
\addcontentsline{toc}{section}{List of Tables}
\listoftables
\vfill
\newpage

\phantomsection
\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\vfill
\newpage

%-----------------------------------------------------------------------------

\phantomsection
\addcontentsline{toc}{section}{Acknowledgments}
\section*{Acknowledgments}

This document represents the work of the people who have served on the C API
Subcommittee of the GraphBLAS Forum.

\vfill
\pagebreak

%-----------------------------------------------------------------------------

\chapter{Introduction}

The GraphBLAS standard defines a set of matrix and vector operations 
based on semiring algebraic structures.  These operations can be used
to express a wide range of graph algorithms.  This document 
defines the mathematical operations required to implement a GraphBLAS 
standard in any programming language.  We refer to 
this as the {\it GraphBLAS Mathematical Specification}.   


The remainder of this document is organized as follows:
\begin{itemize}
%\item Chapter~\ref{Chp:Concepts}: Basic Concepts
%\item Chapter~\ref{Chp:Objects}: Objects
\item Chapter~\ref{Chp:Methods}: Methods
\item Appendix~\ref{Chp:RevHistory}: Revision history
\end{itemize}

%=============================================================================
%=============================================================================

\input{basic_concepts}
\input{objects}

%-----------------------------------------------------------------------------

\chapter{Methods \scott{Rename to Operations}}
\label{Chp:Methods}

This chapter defines the behavior of all the methods in the GraphBLAS C API.
All methods can be declared for use in programs by including the {\tt GraphBLAS.h} header file.

We would like to emphasize that no GraphBLAS method will imply a predefined order over any associative operators. Implementations of the GraphBLAS are encouraged to exploit associativity to optimize performance of any GraphBLAS method. This holds even if the definition of the GraphBLAS method implies a fixed order for the associative operations.

%\input{context_methods}

%\section{Object methods}

This section describes methods that setup and operate on GraphBLAS opaque objects
but are not part of the the GraphBLAS math specification.

%\input{algebra_methods}
%\input{scalar_methods}
%\input{vector_matrix_methods}
%\input{descriptor_methods}

\input{graphblas_operations}
\input{graphblas_computation}

\input{ops_mxm_vxm_mxv}
%\input{ops_ewisemult_ewiseadd}
%\input{ops_extract}
%\input{ops_assign}
%\input{ops_apply}
%\input{ops_select}
%\input{ops_reduce_transpose}
%\input{ops_kronecker}


%=============================================================================
%=============================================================================

%=============================================================================

\appendix
\input{appendices}

\end{document}
